### 2010-03-30 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: fix typos in Section 4, 5.4, 5.5, 6, and 7

In this section, we will fulfil several promises. One promise we made
was to effectively implement the code of the $\Desc$ universe. Another
promise we tacitly was the existence of the type formers for finite
sets. Despite being a perilous pedagogical exercise for the authors,
we hope to convey to the reader the dizzy feeling of levitation,
without the falling.

@@

In Section~\ref{sec:finite-sets}, we have given a specification of
finite sets. It consists of the objects presented in
Figure~\ref{fig:typing-finite-set}. We are going to implement the
$\EnumU$ type former and its constructors. Let us recall their
specification:

@@

We shall now fulfil our promises for $\Desc$. We are going to
implement the codes of the universe of descriptions. Interestingly, in
and by itself, the system of codes, $\Desc$, is nothing but a data-type. We
are in the same situation than with $\EnumU$: we ought to be able to describe
the codes of $\Desc$ in $\Desc$ itself. Hence, this code would be a
first-class citizen, born with the standard, generic equipment of
data-types.

@@

Let us explain how we have proceeded and the obstacle we face. Much as
the data-types we have seen so far, we first have to choose a
constructor. Constructors correspond to the code's constructors:
$\DUnit$, $\DSigma{\!}{\!}$, and $\DIndx{\!}$. The reader will notice that the
tagged notation we have used so far for $\Desc$ constructors now fully
make sense: these were actually the tags we are defining. \note{This
  sentence badly needs a rephrasing.}

@@

In order to capture a notion of higher-order induction, we need to
extend our universe of descriptions. This consists in adding a code
$\DHindx{\!}{\!}$ that takes an indexing set $\V{H}$. The code and its
interpretation are presented in
Figure~\ref{fig:hindx_desc}. Intuitively, $\DHindx{\!}{\!}$ uses the
elements of $H$ to index as many recursive arguments.

@@

Note that $\DIndx{\!}$ is isomorphic to $\DHindx{\Unit}{\!}$. However,
we tolerate this duplication and do not replace $\DIndx{\!}$ by the
more general $\DHindx{\!}{\!}$. Indeed, unlike its counterpart,
$\DIndx{\!}$ is first-order. Therefore, while both codes are
\emph{extensionally} the same, in most practical implementations they
would be dealt with rather differently. While definitional equality
can cope with first order objects, the functional presentation
introduced by $\DHindx{\!}{}$ is unlikely to be amenable to a purely
definitional treatment.

Equipped with $\DHindx{\!}{\!}$, we can describe the extended code of
our universe of data-types, as shown in
Figure~\ref{fig:desc-levitate}.  The $\DUnit$ and $\DIndx{\!}$ cases
remains unchanged, as expected. We successfully describe the
$\DSigma{\!}{\!}$ case, by a simple appeal to the higher-order
induction on $\V{S}$. The $\DHindx{\!}{\!}$ case consists in packing
the $\V{H}$ in $\Set$ with a recursive argument.

@@

\[\stk{
\F{replace} : \stk{\PITEL{\V{D}}{\Desc}
                   \PITEL{\V{X},\V{Y}}{\Set}\\
                   \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                   \All{\V{D}}{\V{X}}{(\LAM{\_}\V{Y})}{\V{xs}} \To
                   \descop{\V{D}}{\V{Y}}} \\
\F{replace}\: \DUnit\:          \V{X}\: \V{Y}\: \Void\:          \Void          \mapsto 
    \Void                                                                                                         \\
\F{replace}\: (\DSigma{\V{S}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{s}}{\V{xs}}{}\: \V{ys}             \mapsto
    \pair{\V{s}}{\F{replace}\: {\V{D}~\V{s}}\: \V{X}\: \V{Y}\: \V{xs}\: \V{ys}}{}                                 \\
\F{replace}\: (\DIndx{\V{D}})\:     \V{X}\: \V{Y}\: \pair{\V{x}}{\V{xs}}{}\: \pair{\V{y}}{\V{ys}}{} \mapsto       \\
\qquad  \pair{\V{y}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{xs}\: \V{ys}}{}                                     \\
\F{replace}\: (\DHindx{\V{H}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{f}}{\V{xs}}{}\: \pair{\V{g}}{\V{ys}}{} \mapsto   \\
\qquad  \pair{\V{g}}{\F{replace}\: \V{D}\{ \V{X}\: \V{Y}\: \V{g}\: \V{ys}}{}
}\]

@@

The astute reader will have been struck by the type of $\F{replace}$:
it is \emph{almost} the morphism part -- sometimes called \emph{map}
-- of the functor $\V{D}$ from $\V{X}$ to $\V{Y}$ in $\Set$. Just as
the $\F{induction}$ is the dependent version of $\cata$, $\F{replace}$
is the dependent version of the map, which uses the induction
hypotheses. For space reason, we will not present the non-dependent
map. It can be found in the Agda model.

@@

We simply add a constructor, $\DVar{\!}$, and define its argument to
be a $\DSigma{\V{X}}{\DUnit}$, that is an element of $\V{X}$. We keep
$\V{E}$ and $\V{D}$ as they were, hence leaving the functor
unchanged. Unfolding the interpretation of this definition, we
convince ourselves that this corresponds to the functor
$\FMFreeMonadD$. The fix-point operation ties the knot and gives us
the full-blown free monad construction.

@@

With the free monad construction, we have seen two kind of generic
operations. Firstly, we have derived a new data-structure from another
one: we make the free monad from its underlying functor. To do so, we
crucially rely on the fact that data-types are nothing but codes. We
are therefore entitled to modify this code and, in this case, extend
it. Extending a data-type might give rise to a more structured object,
as was the case here.  So, secondly, we have equipped this new
data-type with its inherent structure: the \bind\ and
\return\ operations. We have been able to build them as generic
functions.

>>>>

In this section, we will fulfil several promises. One promise we made
was to effectively implement the code of the $\Desc$ universe. Another
promise we tacitly made was the existence of the type formers for
finite sets. Despite being a perilous pedagogical exercise for the
authors, we hope to convey to the reader the dizzy feeling of
levitation, without the falling.

@@

In Section~\ref{sec:finite-sets}, we have given a specification of
finite sets. It consists in the objects presented in
Figure~\ref{fig:typing-finite-set}. We are going to implement the
$\EnumU$ type former and its constructors. Let us recall their
specification:

@@

We shall now fulfil our promises for $\Desc$. We are going to
implement the codes of the universe of descriptions. In and by itself,
the system of codes, $\Desc$, is nothing but a data-type. We are in
the same situation than with $\EnumU$: we ought to be able to describe
the codes of $\Desc$ in $\Desc$ itself. Hence, this code would be a
first-class citizen, born with the standard, generic equipment of
data-types.

@@

Let us explain how we have proceeded and the obstacle we face. Much as
the data-types we have seen so far, we first have to choose a
constructor. Constructors correspond to the code's constructors:
$\DUnit$, $\DSigma{\!}{\!}$, and $\DIndx{\!}$. The reader will notice
that the tagged notation we have used for the $\Desc$ constructors now
fully make sense: these were actually the tags we are
defining. \note{This sentence badly needs a rephrasing.}

@@

In order to capture a notion of higher-order induction, we need to
extend our universe of descriptions. This consists in adding a code
$\DHindx{\!}{\!}$ that takes an indexing set $\V{H}$. The code and its
interpretation are presented in
Figure~\ref{fig:hindx_desc}. Intuitively, $\DHindx{\!}{\!}$ use the
elements of $H$ to index as many recursive arguments.

@@

Note that $\DIndx{\!}$ is isomorphic to $\DHindx{\Unit}{\!}$. However,
we tolerate this duplication and do not replace $\DIndx{\!}$ by the
more general $\DHindx{\!}{\!}$. Indeed, unlike its counterpart,
$\DIndx{\!}$ is first-order. Therefore, while both codes are
\emph{extensionally} the same, in most practical implementation they
would be dealt with rather differently. While definitional equality
can cope with first order objects, the functional presentation
introduced by $\DHindx{\!}{}$ is unlikely to be amenable to a purely
definitional treatment.

Equipped with $\DHindx{\!}{\!}$, we can describe the extended code of
our universe of data-types, as shown in
Figure~\ref{fig:desc-levitate}.  The $\DUnit$ and $\DIndx{\!}$ cases
remain unchanged, as expected. We successfully describe the
$\DSigma{\!}{}$ case, by a simple appeal to the higher-order
induction on $\V{S}$. The $\DHindx{\!}{\!}$ case consists in packing
the $\V{H}$ in $\Set$ with a recursive argument.

@@

\[\stk{
\F{replace} : \stk{\PITEL{\V{D}}{\Desc}
                   \PITEL{\V{X},\V{Y}}{\Set}\\
                   \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                   \All{\V{D}}{\V{X}}{(\LAM{\_}\V{Y})}{\V{xs}} \To
                   \descop{\V{D}}{\V{Y}}} \\
\F{replace}\: \DUnit\:          \V{X}\: \V{Y}\: \Void\:          \Void          \mapsto \Void \\
\F{replace}\: (\DSigma{\V{S}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{s}}{\V{xs}}{}\: \V{ys}             \mapsto
    \pair{\V{s}}{\F{replace}\: {\V{D}~\V{s}}\: \V{X}\: \V{Y}\: \V{xs}\: \V{ys}}{}                                 \\
\F{replace}\: (\DIndx{\V{D}})\:     \V{X}\: \V{Y}\: \pair{\V{x}}{\V{xs}}{}\: \pair{\V{y}}{\V{ys}}{} \mapsto
    \pair{\V{y}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{xs}\: \V{ys}}{}
}\]

@@

@@

We simply add a constructor, $\DVar{\!}$, and define its argument to
be a $\DSigma{\V{X}}{\DUnit}$, that is an element of $\V{X}$. We keep
$\V{E}$ and $\V{D}$ as they were, hence leaving the functor
unchanged. Unfolding the interpretation of this definition, we
convince ourself that this corresponds to the functor
$\FMFreeMonadD$. The fix-point operation ties the knot and gives us
the full-blown free monad construction.

@@

With the free monad construction, we have seen two kinds of generic
operations. Firstly, we have derived a new data-structure from another
one: we make the free monad from its underlying functor. To do so, we
crucially rely on the fact that data-types are nothing but codes. We
are therefore entitled to modify this code and, in this case, extend
it. Extending a data-type might give rise to a more structured object,
as was the case here.  So, secondly, we have equipped this new
data-type with its inherent structure: the \bind\ and
\return\ operations. We have been able to build them as generic
functions.

### 2010-03-30 papers/icfp-2010-desc/paper_discussion.tex
Desc@ICFP: fix typos in Section 4, 5.4, 5.5, 6, and 7

As presented, our type theory suffers from a major weakness. Indeed,
we are subject to Girard's paradox, as we assume that $\Set$ lives in
$\Set$. We made that choice for presentational convenience, as
universe stratification is orthogonal to our work. Nonetheless, our
universe of description rather naturally leads itself to
stratification. Unsurprisingly, $\IDesc{\!}$ at level $l$ is of type
$\Set^{\blue{l+1}}$. Similarly, the interpretation of $\IDesc{\!}$ at
level $l$ is an object of type $\Set^{\blue{l}}$:

@@

Consequently, we can implement the operations and examples developed
above. We refer the reader to our Agda implementation, which take
advantage of set polymorphism to implement the universe of indexed
descriptions at any level. Further, we have coded $\IDesc{\!}$ in
itself and have proved the isomorphism between the host and the
embedded universes.

@@

Another generic programming framework is Scrap Your
Boilerplate~\cite{spj:syb} (SYB). Our proposal is different in various
ways. The corner stone of SYB is the \emph{spine} view of data-type
constructors. A data-type is a spine composed by a constructor applied
to some arguments. Further, this spine is equipped with some
combinators including, primarily, an iterator. In this setting,
generic programs are written by composing these combinators. This
relies on a $\CN{Typeable}$ type-class, allowing dynamic dispatch to
data-type specific operations. As a result, SYB is not reflexive: it
is restricted to data-types instanciating $\CN{Typeable}$. Moreover,
it forbids type-indexed data types: it can only define generic
functions.

@@

The interest in generic programming for dependent types is not new
either. Norell~\cite{norell:msc-thesis} have shown the benefit of
polytypic programming in the setting of Alfa, a precursor of
Agda. Similarly, Verbruggen et al.~\cite{verbruggen:polytype-prog-coq,
  verbruggen:polytype-coq} have developed a framework for polytypic
programming in the Coq theorem prover. However, these works aim at
\emph{modelling} PolyP or Generic Haskell in a dependently-typed
setting, for the purpose of proving correctness properties of Haskell
code. Our approach is different in that we aim at building a
foundation for data-types, in a dependently-typed theory, for a
dependently-typed system.

>>>>

As presented, our type theory suffers from a major weakness. Indeed,
we are subject to Girard's paradox, as we assume that $\Set$ lives in
$\Set$. We made that choice for presentational convenience, as
universe stratification is orthogonal to our work. Nonetheless, our
universes of description rather naturally leads itself to
stratification. Unsurprisingly, $\IDesc{\!}$ at level $l$ is of type
$\Set^{\blue{l+1}}$. Similarly, the interpretation of $\IDesc{\!}$ at
level $l$ is an object of type $\Set^{\blue{l}}$:

@@

The operations and examples developed in this paper stratify just as
well. We refer the reader to our Agda model, which takes advantage of
set polymorphism to implement the universe of indexed descriptions at
any level. Further, we have coded $\IDesc{\!}$ in itself and have
proved the isomorphism between the host and the embedded universes.

@@

Interestingly, our generic induction principle allows a programming
style reminiscent to the Scrap Your Boilerplate~\cite{spj:syb} (SYB)
approach. Like SYB, we provide a generic operator to manipulate and
carry operations over data-types. However, unlike SYB, we do not need
support from the compiler, as required by the $\CN{Typeable}$
class. Indeed, in our setting, data-types are reflected in the
language, hence our ability to manipulate them directly.

@@

The interest in generic programming for dependent types is not new
either. Norell~\cite{norell:msc-thesis} have given a formalization of
polytypic programming in the setting of Alfa, a precursor of
Agda. Similarly, Verbruggen et al.~\cite{verbruggen:polytype-prog-coq,
  verbruggen:polytype-coq} have developed a framework for polytypic
programming in the Coq theorem prover. However, these works aim at
\emph{modelling} PolyP or Generic Haskell in a dependently-typed
setting, for the purpose of proving correctness properties of Haskell
code. Our approach is different in that we aim at building a
foundation for data-types, in a dependently-typed theory, for a
dependently-typed system.
